Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-05-20 paper/ms.tex
update ms

%%%%%%%%%%%%%%%%
% INSTITUTIONS %
%%%%%%%%%%%%%%%%
\newcommand{\caltech}{Department of Astronomy, MC 249-17, California Institute of Technology, Pasadena, CA 91125, USA}
\newcommand{\mitkavli}{MIT Kavli Institute and Department of Physics, 77 Massachusetts Avenue, Cambridge, MA 02139}
\newcommand{\berkeley}{Astronomy Department, University of California, Berkeley, CA 94720, USA}

@@

\author[0000-0002-0514-5538]{Luke~G.~Bouma}
%\altaffiliation{51 Pegasi b Fellow}
\affiliation{\caltech}
\affiliation{51 Pegasi b Fellow}

\author{coauthors, including but not limited to
Hillenbrand, Howard, Isaacson, Palumbo}

% <250 words
\begin{abstract}
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate and lithium content
  can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous work; lithium is
  measured from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which our method is
  applicable.
  %
  We find that our rotation-based ages accurately recover ages of
  known Kepler field clusters; they also agree with
  $\gtrsim$90\% of the independent lithium ages.
  %
  Our resulting catalog includes \nplyounggyrotwosigma\ planets
  younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\ planets
  at 3$\sigma$).
  %
  More broadly, we find that the star formation rate in the Kepler field
  has dropped by a factor of \ratiosfr$\pm$\uncratiosfr\ over the past
  3\,Gyr.  This trend holds true for both planet hosts and the overall
  stellar sample.
  %
  This ``demographic cliff'' is a known feature of the Galaxy's star
  formation history, and it helps clarify the relative balance of
  young and old planets that have been detected to date using
  transits.
  %
\end{abstract}

@@

We measured the lithium equivalent widths using a procedure adapted
from previous work \citep{Bouma_2021}.  Our stars of interest are FGK
stars, and so the continuum in the vicinity of the \ion{Li}{1}
6708\,\AA\ doublet is well-defined.  We Doppler corrected the spectra
to a common reference wavelength by cross-correlating against a high
S/N template for Kepler-1698, chosen because $|\gamma|$$<$10\kms,
$T_{\rm eff}$$\approx$5000\,K and $v\sin i$$\approx$5\,\kms, which
puts it in the middle of our sample's temperature range and gives it
mild line-broadening.  We then trimmed the Doppler-corrected spectra
to a local window centered on the lithium line, using a window width
of 15\,\AA\ (we considered 10\,\AA\ and 20\,\AA; the results were
consistent).  We continuum-normalized by fitting a third-degree
Chebyshev series, while excluding regions with absorption lines.  We
then numerically integrated the resulting spectrum using a
one-component Gaussian with free amplitude, width, and mean, and
estimated uncertainties on the line width through a Monte Carlo
procedure that bootstraped against the local scatter in the continuum.

@@

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.47\textwidth]{li_vs_teff_koi_X_JUMP_eagles.pdf}
	\end{center}
	\vspace{-0.6cm}
	\caption{
    {\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
    for planet-hosting stars.} These measurements were made from
    Keck/HIRES spectra collected between 2009 and 2024 that comprised
    \nhireshours\,hours of open-shutter time.  Lines are the ``mean''
    isochrones from the same models as in Figure~\ref{fig:models}.  The
    intrinsic dispersion beyond $\approx$1\,Gyr becomes much larger
    than changes in the mean.  Some stars with lithium detections do
    not have detected rotation periods, and vice-versa.
		\label{fig:li_vs_teff}
	}
\end{figure}

@@

Stars that formed in the same birth cluster, at approximately the same time,
are the gold standard for the astronomical age scale
\citep{Soderblom10}.
Before Gaia, only a handful of $\gtrsim$1\,Gyr open clusters
were known in the Kepler field.

The gold standard for stellar ages comes from open clusters

Figure~\ref{fig:agescalecompone} shows the result of a literature
comparison against ages reported by \citet{Berger_2020a_catalog},
\citet{2021AJ....161..189L}, \citet{2023ApJ...952..131M}, and
\citet{2024AJ....167..159L}.
We followed any guidance available from these respective studies for
removing ages that were flagged as unreliable.
Rather than comparing the ages of the stars directly against each other,
we compared them against a common benchmark: open clusters.

Cluster memberships are a delicate subject.
Figure~\ref{fig:agescalecompone} is showing putative ages for
a set of spatially and kinematically selected candidate cluster members.  For NGC6811 and NGC6819, we
adopted candidate members from
\citet{2018A&A...618A..93C,CantatGaudin_2020} and
\citet{Kounkel_2020}. 
For Melange-3 we used the candidate members reported by
\citet{2022AJ....164...88B} and required ``offset'' tangential
velocities below 2\kms.  For Cep-Her we
used candidate members from Table~2 of \citet{Bouma_2022b}, requiring
``weights'' exceeding 0.1.
These lists almost certainly include non-members, in fractions that we
do not aim to quantify.  Even with these contaminants, we can compare
the relative age distributions derived by the different studies.

There are two main metrics for success in this test: the first is
whether the reported ages agree with the cluster age (accuracy).  The
second is whether the reported age precisions agree with their
dispersion around the cluster age.

Starting with the isochrone ages from \citet{Berger_2020a_catalog},
the results should not be particularly surprising.
The isochrone ages are most accurate and precise (in relative terms)
for the oldest cluster, NGC6819.
For the 1\,Gyr NGC6811, about half the stars have ages overestimated
by a factor of three.
For the youngest clusters, the disagreement is greater.
The uncertainties seem accurate at 2.5\,Gyr, but perhaps less so at
$\lesssim$1\,Gyr.
An important virtue of these isochrone ages is that they are
calculated for far more stars in the two largest clusters than the
other methods.
They are calculable for far more stars than rotation-based ages.

The ``gyro-kinematic'' ages from \citet{2021AJ....161..189L} do...
{\bf todo explain}

The rotation-only ages calculated using the model described by
\citet{2024AJ....167..159L} have some peculiarities compared against
cluster ages.
While NGC6811 seems to be in general agreement, NGC6819 has an
abnormally large population of young reported systems, with apparently
precise ages.  All the younger cluster members ($<$300\,Myr) similarly
have reported ages that are biased old, with narrow uncertainties.

\citet{2023ApJ...952..131M} looks great.  {\bf todo explain method}.
However it still fails for the youngest clusters.

@@

In this work, we were interested in not only the overall age
distribution of the Kepler field, but also in the reliability of
individual ages for young stars known to host planets.  The quality
bitmask scheme in Table~\ref{tab:planets} provided one path toward
this end.  Our other method was to compare the cases for which lithium
and rotation overlap in sensitivity -- G and K dwarfs with
$t$$\lesssim$1\,Gyr.  Our results for such stars include
\ltonegyrhighqconfirmedtwosided\ planets orbiting ``high-quality''
stars for which the two age indicators are both constraining and in
agreement (stars with $Q_{\rm star}$ bits zero through nine not
raised, and no planet quality flags raised).  Conversely, we found
only one system, Kepler-786, with a precise lithium age
(\kepseveneightsix\,Myr), but no rotation signal.

@@

A broader question, beyond the scope of this work, concerns the ages
of the older $\gtrsim$3\,Gyr stars that we do not treat.  For
age-dating models based only on rotation rates, or else on a
combination of rotation and spin-modulation amplitude
\citep{2023ApJ...952..131M}, an important challenge is that the
detectability of the rotation signals for 3-10\,Gyr stars, especially
Sun-like stars, is at the limits of the current data
\citep{2022ApJ...937...94M}.  In the ``old star'' regime, methods
using either asteroseismology
\citep{vanSaders_2016,2024ApJ...962..138S}, or else using both the
evolution of stellar luminosity and stellar rotation
\citep{Angus_2019} seem the most capable of providing useful age
constraints for single stars.  Incorporating kinematics
\citep{2021AJ....161..189L} seems likely to remain most useful for
stellar ensembles, due to the stochastically of kinematic heating.

@@

\section{Cases With ``Inconsistent'' Rotation and Lithium-Based Ages}
\label{app:inconsistent}

@@

\begin{enumerate}
  \item Compare Mathur2023 ages against my ages, and (more interestingly) against the
    cluster scale using NGC6811, Theia520, Melange3, and CepHer stars.
    Could compare against asteroseis too.
    %
  \item Intro needs better discussion of Mathur2023 and Lu2021
    %
  \item Consider moving the bit list description to its own appendix.
    %
  \item Rerun gyro calculation with denser grid, fixing quantization
    and extending upper limit to 5gyr.  (Will use updated Teff
    uncertainties)
\end{enumerate}

>>>>

%%%%%%%%%%%%%%%%
% INSTITUTIONS %
%%%%%%%%%%%%%%%%
\newcommand{\caltech}{Department of Astronomy, MC 249-17, California Institute of Technology, Pasadena, CA 91125, USA}
\newcommand{\mitkavli}{MIT Kavli Institute and Department of Physics,
77 Massachusetts Avenue, Cambridge, MA 02139, USA}
\newcommand{\berkeley}{Astronomy Department, University of California,
Berkeley, CA 94720, USA}
\newcommand{\cmu}{Department of Statistics \& Data Science, Carnegie
Mellon University, Pittsburgh, PA 15213, USA}

@@

\author[0000-0002-0514-5538]{Luke~G.~Bouma}
\altaffiliation{51 Pegasi b Fellow}
\affiliation{\caltech}
%\affiliation{51 Pegasi b Fellow}

\author{and coauthors, including but not limited to...}

\author{Lynne~A.~Hillenbrand}
\affiliation{\caltech}

\author[0000-0001-8638-0320]{Andrew W. Howard}
\affiliation{\caltech}

\author[0000-0002-0531-1073]{Howard Isaacson}
\affiliation{\berkeley}

\author[0000-0001-7967-1795]{Elsa~K.~Palumbo}
\affiliation{\cmu}

% <250 words
\begin{abstract}
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate and lithium content
  can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous work; lithium is
  measured from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which our method is
  applicable.
  %
  We find that our rotation-based ages accurately recover ages of
  stars in open clusters; they also agree with
  $\gtrsim$90\% of the independent lithium ages.
  %
  The results include \nplyounggyrotwosigma\ planets
  younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\ planets
  at 3$\sigma$).
  %
  More broadly, we find that the star formation rate in the Kepler field
  has dropped by a factor of \ratiosfr$\pm$\uncratiosfr\ over the past
  3\,Gyr.  This trend holds for both planet hosts and for the overall
  stellar sample.
  %
  This ``demographic cliff'' is a known feature of the Galaxy's star
  formation history, and it helps clarify the relative balance of
  young and old planets that have been detected to date using
  transits.
  %
\end{abstract}

@@

We measured the lithium equivalent widths using a procedure adapted
from previous work \citep{Bouma_2021}.  Our stars of interest are FGK
stars, and so the continuum in the vicinity of the \ion{Li}{1}
6708\,\AA\ doublet is well-defined.  We Doppler corrected the spectra
to a common reference wavelength by cross-correlating against a high
S/N template for Kepler-1698, chosen because $|\gamma|$$<$10\kms,
$T_{\rm eff}$$\approx$5000\,K and $v\sin i$$\approx$5\,\kms, which
puts it in the middle of our sample's temperature range and gives it
mild line-broadening.  We then trimmed the Doppler-corrected spectra
to a local window centered on the lithium line, using a window width
of 15\,\AA\ (we also considered 10\,\AA\ and 20\,\AA; the results were
consistent).  We continuum-normalized by fitting a third-degree
Chebyshev series, while excluding regions with absorption lines.  We
then numerically integrated the resulting spectrum using a
one-component Gaussian with free amplitude, width, and mean, and
estimated uncertainties on the line width through a Monte Carlo
procedure that bootstrapped against the local scatter in the continuum.

@@

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.47\textwidth]{li_vs_teff_koi_X_JUMP_eagles.pdf}
	\end{center}
	\vspace{-0.6cm}
	\caption{
    {\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
    for planet-hosting stars.} These measurements were made from
    Keck/HIRES spectra collected between 2009 and 2024.  Lines are the ``mean''
    isochrones from the same models as in Figure~\ref{fig:models}.  The
    intrinsic dispersion at $\gtrsim$1\,Gyr becomes much larger
    than changes in the mean.  Some stars with lithium detections do
    not have detected rotation periods, and vice-versa.
		\label{fig:li_vs_teff}
	}
\end{figure}

@@

Stars that formed in the same birth cluster 
are the gold standard for the astronomical age scale
\citep{Soderblom_2010}.
Before Gaia, the few $\gtrsim$1\,Gyr open clusters
known in the Kepler field had been visually identified by
\citet{1864RSPT..154....1H}.
Gaia proper motions and distances 
enabled the discovery of stellar ensembles that are typically more diffuse,
but which nonetheless
share a common age 
based on e.g.~isochrone, rotation, and lithium dating
\cite[e.g.][]{2019AJ....158..122K,Bouma_2022b,Barber_2022}.

In Figure~\ref{fig:agescalecompone}, we compare 
ages from the literature against the ages of stars in the richest
clusters in the Kepler field.
In particular, we drew ages for Kepler stars from \citet{Berger_2020a_catalog},
\citet{2021AJ....161..189L}, \citet{2023ApJ...952..131M}, and
\citet{2024AJ....167..159L}.
We followed any guidance available from these respective studies for
removing ages that were explicitly flagged as unreliable.

Cluster memberships are a delicate subject.
Figure~\ref{fig:agescalecompone} is showing reported ages for
a set of spatially and kinematically selected set of stars that could be cluster members, or
they could be field interlopers.  For NGC6811 and NGC6819, we
adopted candidate members from
\citet{2018A&A...618A..93C,CantatGaudin_2020} and
\citet{Kounkel_2020}. 
For Melange-3 we used the candidates reported by
\citet{Barber_2022} and required ``offset'' tangential
velocities below 2\,\kms.  For Cep-Her we
used candidate members from Table~2 of \citet{Bouma_2022b}, requiring
``weights'' exceeding 0.1.
Even with contaminants, we can compare
the relative age distributions derived by the different studies because in all
instances we are comparing reported ages against a fixed list of stars.

There are two main metrics for success in this test: accuracy
in whether the reported ages agree with the cluster age, and precision in
whether the reported age uncertainties agree with their
dispersion around the cluster age.

The point of Figure~\ref{fig:agescalecompone} is hopefully clear.
While previous studies have been able to derive accurate ages for
stars $\gtrsim$1\,Gyr old, the ages of sub-gigayear stars have historically been
overestimated by anywhere from 0.5--2\,dex.
The situation is particularly severe for the 40\,Myr stars in Cep-Her,
but it extends out to 150--300\,Myr stars in Melange-3 and Theia~520.
For isochrone and kinematic ages,
this is not surprising because these methods use stellar parameters that do
not appreciably change at $t$$\lesssim$1\,Gyr.
For the rotation-based ages quoted by
\citet{2023ApJ...952..131M} and \citet{2024AJ....167..159L}, the discrepancy
between their ages and the cluster age scale is likely caused by details of the calibration 
methodologies.
The physical spin-down model used by
\citet{2023ApJ...952..131M},
and the gyro-kinematic calibration ages used by \citet{2024AJ....167..159L},
may not reproduce
the rotation sequences of open clusters such as Praesepe, the Hyades, and the Pleiades.
The statistical age uncertainties yielded by these approaches
similarly may not marginalize over the intrinsic
scatter in rotation periods present in stars of a fixed mass at any given age.

%The rotation-based ages that we have calculated have a precision
%limited by the intrinsic scatter in the cluster rotation sequences.
%Within the age range accessible through the clusters, they seem to be
%mostly accurate,
%with the one exception being that our ages for stars in Cep-Her 
%have a typical median of $\approx$80\,Myr, rather than 40\,Myr.
%The reason is that 80\,Myr is the lower bound of 
%the \texttt{gyro-interp}  age grid; pre-main-sequence
%spin-up is outside of our calibration regime.

@@

We were interested in not only the overall age
distribution of the Kepler field, but also in the reliability of
individual ages for young stars known to host planets.  
We assessed this reliability using
quality
bitmasks (Table~\ref{tab:planets}), 
a comparison against open cluster ages (Section~\ref{sec:litagecomparison}),
and a comparison between rotation and lithium ages.
The open cluster comparison suggested that our ages were generally
accurate at $\ll$0.5\,dex across 0.04--2.5\,Gyr, comparable to their
stated precision level.
Nonetheless,
outliers do exist in the upper-left panel of Figure~\ref{fig:agescalecompone},
typically at most 10\% of the population in each cluster.
These stars could be either field star contaminants, or else anomalous
stars whose physical rotation histories were altered by processes not
captured by our statistical uncertainties.
Similarly, although
we found $t_{\rm gyro}$ and $t_{\rm Li}$ to agree
for 
\ltonegyrhighqconfirmedtwosided\ planets orbiting ``high-quality''
stars, we did find
one system, Kepler-786, with a precise lithium age
(\kepseveneightsix\,Myr), but no rotation signal.

@@

A broader question, beyond the scope of this work, concerns the ages
of stars $\gtrsim$4\,Gyr old.  For
age-dating models based only on rotation rates, or else on a
combination of rotation and spin-modulation amplitude
\citep{2023ApJ...952..131M}, an important challenge is that the
detectability of the rotation signals for 3-10\,Gyr stars, especially
Sun-like stars, is at the limits of the current data
\citep{2022ApJ...937...94M}.  In the ``old star'' regime, methods
using either asteroseismology
\citep{vanSaders_2016,2024ApJ...962..138S}, or else using both the
evolution of stellar luminosity and stellar rotation
\citep{Angus_2019} seem the most capable of providing useful age
constraints for single stars.  Incorporating kinematics
\citep{2021AJ....161..189L} seems likely to remain most useful for
stellar ensembles, due to the stochasticity of kinematic heating.

@@

\appendix
\section{Cases With ``Inconsistent'' Rotation and Lithium-Based Ages}
\label{app:inconsistent}

@@

\begin{enumerate}
%  \item Compare Mathur2023 ages against my ages, and (more interestingly) against the
%    cluster scale using NGC6811, Theia520, Melange3, and CepHer stars.
%    Could compare against asteroseis too.
%    %
%  \item Intro needs better discussion of Mathur2023 and Lu2021
    %
  \item Consider moving the bit list description to its own appendix.
    %
  \item Rerun gyro calculation with denser grid, fixing quantization
    and extending upper limit to 5gyr.  (Will use updated Teff
    uncertainties)
\end{enumerate}
